Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2015-06-19 tmm.tex
add data table for model verification

We apply the above linear decoding model to the case of quality scalability of SVC and get a linear error model. According to (\ref{eq:linear}), for a bitstream just containing the base layer, we have:
\begin{equation}
\label{eq:base0}
s_{B} = {\rm\bf M}s_{B} + {\rm\bf T}_{B}c_{B} + p \: ,
\end{equation}
where the subscript $B$ indicates the base layer variables. In quality scalability, the enhancement is mainly the refinement of transform coefficients, so for a bitstream containing an enhancement packet set $I_1$, the reconstruction relationship can be expressed as:
\begin{equation}
\label{eq:subset1}
s_{I_1} = {\rm\bf M}s_{I_1} + {\rm\bf T}_{B} c_{B} + \Big( \sum_{i \in I_1}{\rm\bf T}_i c_i \Big) + p \: ,
\end{equation}
where $i$ refers to an enhancement packet as an element of $I_1$. $c_i$ is a vector containing coefficients within the enhancement packet $i$. ${\rm\bf T_i}$ is the transform matrix depending on the QP value of $i$. ${\rm\bf T}_i c_i$ gives the residual sample values decoded from $i$.

@@

The above model is called Linear Error Model (LEM). Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}) the value of M and p remain the same, since in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction. This argument no longer holds for the other scalable dimensions (e.g., spatial scalability), so the model does not simply apply for those cases. Only quality scalable videos are used in our system, and further research for other scalable dimensions can be our future work.

@@

To verify the linear error model, we conduct an experiment where random combinations of enhancement packets are removed from a video bitstream, and the distortion estimated by the linear error model are compared with the actual distortion caused by the removal. Experimental result of the sequence {\em Foreman} is shown in Fig. \ref{fig:model_verification}. The bitstream has 8-frame GOPs and three quality layers, the same as illustrated in Fig. \ref{fig:GOP_Structure}. There are 33 frames in the sequence. The horizontal axis stands for frame index, and the vertical axis for MSE measurement. The solid line represents the actual MSE of a frame, and the dot line shows the estimated MSE using our Linear Error Model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of the proposed model. The distortion estimation error is less than 5\% in our experiments.

\begin{figure}[h]
\centering
\includegraphics[width = 0.9\linewidth]{ModelVerification.png}
\caption{Result of SVC Linear Error Model verification \label{fig:model_verification}}
\end{figure}

@@

In this subsection, we utilize the Linear Error Model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction algorithm uses a greedy method \cite{GreedyAlgo} to get a sub-optimal solution. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi_{i} = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

In some scenarios, the priority assignment and bitstream extraction needs to be conducted without access to the original YUV sequence. For example, if the bitstream extraction is not done at the encoding side, it's most likely that the original sequence is not available. This would make the bitstream extraction a more difficult problem, since the distortion is harder to obtain for lacking of reference signal. Comparing to the problem of SVC bitstream extraction with reference, this kind of ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice. In this subsection, we handle this no-reference problem as well and modify the proposed algorithm to meet the special characteristic caused by lacking of reference.

In the no-reference case, since we do not have access to the original sequence, we can not get the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used above. A straightforward idea is to set $e_{full}$ to zero, meaning we calculate distortion with respect to the fully reconstructed signal, rather than to the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. We find that, since discarding some enhancement packets will not bring much significant error relative to the fully reconstructed signal, the calculated PSNR values are too large to well represent the real distortion. To address this issue, instead of calculating PSNR, we directly use MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be a simple and effective solution, and the extraction performance turns out to be much better.

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different data rates. Next, we propose the quality control algorithm based on the PID control mechanism, which is responsible for determining a suitable data rate according to current bandwidth condition.

>>>>

We apply the above linear decoding model to the case of quality scalability of SVC and get a Linear Error Model (LEM). According to (\ref{eq:linear}), for a bitstream just containing the base layer, we have:
\begin{equation}
\label{eq:base0}
s_{B} = {\rm\bf M}s_{B} + {\rm\bf T}_{B}c_{B} + p \: ,
\end{equation}
where the subscript $B$ indicates the base layer variables. In quality scalability, the enhancement is mainly the refinement of transform coefficients, so for a bitstream containing an enhancement packet set $I_1$, the reconstruction relationship can be expressed as:
\begin{equation}
\label{eq:subset1}
s_{I_1} = {\rm\bf M}s_{I_1} + {\rm\bf T}_{B} c_{B} + \Big( \sum_{i \in I_1}{\rm\bf T}_i c_i \Big) + p \: ,
\end{equation}
where $i$ refers to an enhancement packet as an element of $I_1$. $c_i$ is a vector containing coefficients within the enhancement packet $i$. ${\rm\bf T_i}$ is the transform matrix depending on the QP value of $i$. ${\rm\bf T}_i c_i$ gives the residual sample values decoded from $i$.

@@

Note that in equations (\ref{eq:base0})-(\ref{eq:subset2}) the value of M and p remain the same, since in the case of quality scalability the enhancement packets are generated only by manipulating the transform coefficients, and not interfered with the motion compensation and static prediction. This argument no longer holds for the other scalable dimensions (e.g., spatial scalability), so LEM does not simply apply for those cases. Therefor, only quality scalable videos are used in our system, and further research for other scalable dimensions can be our future work.

@@

To verify the linear error model, we conduct an experiment where random combinations of enhancement packets are removed from a video bitstream, and the distortion estimated by the linear error model are compared with the actual distortion caused by the removal. The video stream has 33 frames, and the same GOP structure and quality layers as illustrated in Fig. \ref{fig:GOP_Structure}. One comparison result for the sequence {\em Foreman} is shown in Fig. \ref{fig:model_verification}. The horizontal axis stands for frame index, and the vertical axis for MSE measurement. The solid line represents the actual MSE of a frame, and the dot line shows the estimated MSE using our Linear Error Model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of the proposed model. Table \ref{tab:estimation-error} lists the distortion estimation error for different sequences and shows that the average error is only about 5\%.

\begin{figure}[h]
\centering
\includegraphics[width = 0.9\linewidth]{ModelVerification.png}
\caption{Comparison of actual and estimated distortion for sequence {\em Foreman} \label{fig:model_verification}}
\end{figure}

\begin{table*}
\centering
\caption{Performance of distortion estimation using LEM}
\label{tab:estimation-error}
\begin{tabular}{c|*{8}{p{1.2cm}<{\centering}|}{p{1.2cm}<{\centering}}}
	\hline\hline
	Sequence name & {\em Bus} & {\em City} & {\em Crew} & {\em Football} & {\em Foreman} & {\em Harbour} & {\em Mobile} & {\em Soccer} & \textbf{Average} \\ \hline
	Estimation error  & 2.83\% & 2.39\% & 10.36\% & 7.08\% & 4.95\% & 2.63\% & 5.77\% & 5.26\% & \textbf{5.16\%} \\ \hline
\end{tabular}
\end{table*}

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction algorithm uses a greedy method \cite{GreedyAlgo} to get a sub-optimal solution. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi_{i} = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

@@

In some scenarios, the priority assignment and bitstream extraction needs to be conducted without access to the original YUV sequence. For example, if the bitstream extraction is not done at the encoding side, it's most likely that the original sequence is not available. This would make the bitstream extraction a more difficult problem, since the distortion is harder to obtain for lacking of reference signal. Comparing to the problem of SVC bitstream extraction with reference, this kind of ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice. In this subsection, we consider this no-reference problem and modify the proposed algorithm to meet the special characteristic caused by lacking of reference.

In the no-reference case, since we do not have access to the original sequence, we can not get the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used above. A straightforward idea is to set $e_{full}$ to zero, meaning we calculate distortion with respect to the fully reconstructed signal, rather than to the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. We find that, since discarding some enhancement packets will not bring much significant error relative to the fully reconstructed signal, the calculated PSNR values are too large to well represent the real distortion. To address this issue, instead of calculating PSNR, we directly use MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be an effective solution, and the extraction performance turns out to be much better.

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different data rates. Next, we propose the quality control algorithm based on the PID mechanism, which is responsible for determining a suitable data rate according to current bandwidth condition.
